A Monte Carlo Synthetic-Acceleration Method for Solving the Thermal Radiation Diffusion Equation\tnoteref{ornl-cr}

### 2013-02-19 doc/jcomp2012/jcomp2012.tex
fixing document build

A timing and convergence study is used to demonstrate the
effectiveness of the adjoint method as compared to the direct
method. To assess both the CPU time and number of iterations required
to converge to a solution, a problem of constant $\Delta$ was used
with varying values of grid size, fixing the spectral radius of the
system at a constant value for each variation. Both the five-point and
nine-point stencils were used with both the direct and adjoint
solvers. For each case, $N \times N$ total random walk permutations
were computed per MCSA iteration where $N \times N$ is the number of
discrete grid points in the system. Solver parameters were set to a
weight cutoff of \sn{1}{-4} for the stochastic linear solver and a
convergence tolerance of \sn{1}{-8} for the MCSA iterative solver. All
computations presented in this section were completed on a 2.66 GHz
Intel Core i7 machine with 8 GB 1067 MHz DDR3
memory. Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed
for each case to converge in seconds, and
Figure~\ref{fig:poisson_iterations} gives the number of iterations
needed for each case to converge relative to the problem size.
\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{dir_adj_cpu.pdf}
  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
    $N \times N$ square mesh). Both the adjoint and direct solvers are
    used with the five point and nine point stencils. A CPU time
    speedup is noted with the adjoint method due to the higher density
    of random walk events in regions with a large residual.}
  \label{fig:poisson_cpu_time}
\end{figure}

\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{dir_adj_iterations.pdf}
  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
    \times N$ square mesh). Both the adjoint and direct solvers are
    used with the five-point and nine-point stencils. }
  \label{fig:poisson_iterations}
\end{figure}

@@

As an additional comparison, the convergence behavior of MCSA can be
analyzed using both the adjoint and direct solvers to detect any
performance benefits. To assess the convergence properties of MCSA
using each solver and stencil, the infinity norm of the residual
computed in Eq.~(\ref{eq:MCSA-residual}) was collected at each
iteration for a fixed problem size of
$N=500$. Figure~\ref{fig:poisson_convergence} gives the results of
these computations. First, it is worthy to note on the semilog plot
that we are indeed achieving the expected exponential convergence from
MCSA with both Monte Carlo solvers. Second, we note that using the
adjoint method with the same number of stochastic histores per MCSA
iteration gives a faster rate of converge for the same reasons as
above.
\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{dir_adj_convergence.pdf}
  \caption{\sl Infinity norm of the solution residual vs. iteration
    number for a problem of size $N=500$. Both the adjoint and direct
    solvers are used with the five point and nine point stencils. A
    higher rate of convergence is observed for MCSA using the adjoint
    Monte Carlo solver as compared to the direct method when both
    solvers compute the same number of random walks per iteration.}
  \label{fig:poisson_convergence}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{seq_cpu.pdf}
  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
    $N \times N$ square mesh). Both the Sequential Monte Carlo and
    MCSA solvers are used with the five point stencils and the adjoint
    Monte Carlo solver. The number of random walks was twice the
    number of discrete states in the system in order to ensure
    convergence in the Sequential Monte Carlo method.}
  \label{fig:seq_cpu_time}
\end{figure}

\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{seq_iterations.pdf}
  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
    \times N$ square mesh). Both the Sequential Monte Carlo and MCSA
    solvers are used with the five point stencils and the adjoint
    Monte Carlo solver.}
  \label{fig:seq_iterations}
\end{figure}

The benefits of using a synthetic acceleration scheme are also noted
when the infinity norm of the residual computed computed at each
iteration for both methods was collected at each iteration for a fixed
problem sizes of $N=100$ and $N=500$ as shown in figures
Figure~\ref{fig:seq_100} and \ref{fig:seq_500} respectively. In both
cases, the Sequential method is subject to two regimes of exponential
convergence with the later regime converging the slowest while the
MCSA method exhibits a single rate of exponential convergence, much
higher than that computed by Halton's method. Even with the doubling
of the number of stochastic histories computed per time step in order
to ensure convergence for the Sequential method, we still see
robustness issues with a non-monotonically decreasing residual
observed in both cases and more pronounced for the $N=100$ case. In
both cases the MCSA solver is observed to be robust with a
monotonically decreasing residual.

\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{seq_conv_100.pdf}
  \caption{\sl Infinity norm of the solution residual vs. iteration
    number for a problem of size $N=100$. Both the Sequential Monte
    Carlo and MCSA solvers are used with the five point stencils and
    the adjoint Monte Carlo solver.}
  \label{fig:seq_100}
\end{figure}

\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{seq_conv_500.pdf}
  \caption{\sl Infinity norm of the solution residual vs. iteration
    number for a problem of size $N=500$. Both the Sequential Monte
    Carlo and MCSA solvers are used with the five point stencils and
    the adjoint Monte Carlo solver.}
  \label{fig:seq_500}
\end{figure}

@@

Applying backward-Euler time differencing to Eq.~(\ref{eq:eq-diff}) and
lagging the temperature-dependent coefficients at $t^n$ gives
\begin{equation}
  -\nabla\cdot \Dn\grad{\phi} + \sign\phi = \qn\:,
\end{equation}
where we have suppressed the $n+1$ indices on $\phi$ and
\begin{align}
  \sign &= \frac{\rho \Cv^n}{4ac(\Tn)^3\dt} + \frac{1}{c\dt}\:,\\ \qn
  &= \sign\phin + Q^n\:,\\ \phin &= ac(\Tn)^4\:.
\end{align}
Applying Fick's Law,
\begin{equation}
  \ve{F} = -D\grad{\phi}\:,
  \label{eq:fick}
\end{equation}
where $\ve{F}$ is the radiation flux, we define a flux-balance equation,
\begin{equation}
  \dive{F} + \sign\phi = \qn\:.
  \label{eq:flux-balance}
\end{equation}
Integrating Eq.~(\ref{eq:flux-balance}) over the cell illustrated in
Figure~\ref{fig:cell} gives
\begin{equation}
  \iiint\dive{F}\,d\ve{r} +
  \sign_{ijk}\phi_{ijk}V_{ijk}=\qn_{ijk}V_{ijk}\:,
  \label{eq:flux-balance-vol-integral}
\end{equation}
and $V_{ijk} = \Delta_i\Delta_j\Delta_k$.
\begin{figure}[h!]
  \centerline{ \includegraphics[clip]{mesh_cell.pdf} }
  \caption{Orthogonal 3D mesh cell.}
  \label{fig:cell}
\end{figure}
We now apply the divergence theorem,
\begin{equation}
  \begin{split}
    \iiint\dive{F}\,d\ve{r} &= \oint\hat{\ve{n}}\cdot\ve{F}\,dA\\ &=
    \sum_{f=1}^{6}\hat{\ve{n}}_f\cdot\ve{F}_f A_f\:,
  \end{split}
\end{equation}
to Eq.(\ref{eq:flux-balance-vol-integral}) yielding
\begin{equation}
  \begin{split}
    (F_{i+1/2} - F_{i-1/2})\Dj\Dk &+ (F_{j+1/2} - F_{j-1/2})\Di\Dk +
    (F_{k+1/2}\\ &- F_{k-1/2})\Di\Dj + \sign_{ijk}\phi_{ijk}V_{ijk} =
    \qn_{ijk}V_{ijk}\:.
  \end{split}
  \label{eq:flux-balance-difference}
\end{equation}

@@

The connectivity of an arbitrary cell is illustrated in the 3D mesh segment in
Figure~\ref{fig:mesh_segment}.  For a particle residing in the blue cell, any
collision results in a scatter to one of the six adjacent cells since
$p_{ii}=0$ in the preconditioned system.  We can easily calculate the
probabilities for each scattering event from
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) without explicitly forming
the iteration matrix.  In this mesh cell, the probabilities for leakage out of
each face are derived from the equations from each adjacent cell.  Using
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) we can explicitly write the
equations from each adjacent cell, here only noting the contribution from cell
7:
\begin{align*}
  \phi_1 &= \frac{\sigm_{7\rightarrow 1}}{\sigma_{T\,1}}
  \frac{\Delta_7}{\Delta_1}\phi_7 + \ldots\:,\\ 
  %%
  \phi_2 &= \frac{\sigp_{7\rightarrow 2}}
  {\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2}\phi_7 + \ldots\:,\\
  %%
  \phi_3 &= \frac{\sigm_{7\rightarrow 3}}{\sigma_{T\,3}}
  \frac{\Delta_7}{\Delta_3}\phi_7 + \ldots\:,\\ 
  %%
  \phi_4 &= \frac{\sigp_{7\rightarrow 4}}{\sigma_{T\,4}}
  \frac{\Delta_7}{\Delta_4}\phi_7 + \ldots\:,\\
  %%
  \phi_5 &= \frac{\sigm_{7\rightarrow 5}}{\sigma_{T\,5}}
  \frac{\Delta_7}{\Delta_5}\phi_7 + \ldots\:,\\
  %%
  \phi_6 &= \frac{\sigp_{7\rightarrow 6}}{\sigma_{T\,6}}
  \frac{\Delta_7}{\Delta_6}\phi_7 + \ldots\:.\\ 
\end{align*} 
A quick view of these equations reveals that the only non-zero entries in
column 7 of the iteration matrix are
\begin{equation*}
  \begin{pmatrix}
    h_{1,7} & h_{2,7} & h_{3,7} & h_{4,7} & h_{5,7} & h_{6,7}
  \end{pmatrix}^{T}
  =
  \begin{pmatrix}
    \frac{\sigm_{7\rightarrow
        1}}{\sigma_{T\,1}}\frac{\Delta_7}{\Delta_1}
    \\ \frac{\sigp_{7\rightarrow
        2}}{\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2}
    \\ \frac{\sigm_{7\rightarrow
        3}}{\sigma_{T\,3}}\frac{\Delta_7}{\Delta_3}
    \\ \frac{\sigp_{7\rightarrow
        4}}{\sigma_{T\,4}}\frac{\Delta_7}{\Delta_4}
    \\ \frac{\sigm_{7\rightarrow
        5}}{\sigma_{T\,5}}\frac{\Delta_7}{\Delta_5}
    \\ \frac{\sigp_{7\rightarrow
        6}}{\sigma_{T\,6}}\frac{\Delta_7}{\Delta_6}
  \end{pmatrix}\:,
\end{equation*}
where, following Eq.~(\ref{eq:point_iteration}), $\ve{H} = (\vI -
\ve{M}^{-1}\ve{D})$ is the iteration matrix.  Now, using
Eq.~(\ref{eq:adjoint-probability}), we define the scattering probabilities for
cell 7 as
\begin{equation*}
  P_{7\rightarrow n} = p_{7,n} = \frac{|h_{n,7}|}{|h_{1,7} + h_{2,7} +
    h_{3,7} + h_{4,7} + h_{5,7} + h_{6,7}|}\:,\quad n\in
  \{1,\ldots,6\}\:.
\end{equation*}
The weight change for each scattering event is calculated using
Eq.~(\ref{eq:adjoint-weight}):
\begin{equation*}
  W_{m} = \frac{h_{n,7}}{p_{7,n}}W_{m-1}\:,\quad n\in
  \{1,\ldots,6\}\:,
\end{equation*}
where $m$ is the step-index in the particle history.  We also note that, for
the 3D Cartesian mesh considered in this work, the maximum number of non-zero
entries in columns of the iteration matrix is 6, generating a sparse system.
Cells along the problem boundary can have fewer than six entries.
\begin{figure}[h!]
  \centerline{ \includegraphics[clip]{3D_mesh_segment.pdf}}
  \caption{3D mesh-segment.  The particle's current cell (in blue) is
    connected to 6 adjacent cells.  The cell numbers are plotted
    alongside the mesh such that the blue cell has index 7, the high
    $z$ cell has index 2, etc.}
  \label{fig:mesh_segment}
\end{figure}

@@

Figure~\ref{fig:marshak_10} shows the analytic Marshak Wave solution compared
to the MCSA solution.  As opposed to most Monte Carlo techniques, the MCSA
method generates solutions that are numerically equivalent to the PCG and PFIX
results.  Because the MCSA method converges the residual of the system, the
numerical accuracy is determined by the stopping criterion shown in
Eq.~(\ref{eq:stopping-criteria}).  Also, the MCSA method is not bound by the
Central Limit Theorem so convergence is not restricted to
$1/\sqrt{N}$~\cite{halton_1994,evans_2003}.
\begin{figure}[h!]
  \centerline{ \includegraphics[width=5in,clip]{marshak_10.pdf}}
  \caption{ Solution of the Marshak wave problem at $t=100$~ns.  The analytic
    solution is compared to the MCSA solution.  The MCSA, PCG, and PFIXED
    methods are all converged to the same tolerance and therefore give
    numerically equivalent answers.}
  \label{fig:marshak_10}
\end{figure}

As stated in \S~\ref{sec:monte-carlo-matrix}, the MCSA method will converge
when $\rho(\ve{H})< 1$, where $\ve{H}$ is the iteration matrix and is defined
$\ve{H} = (\vI - \ve{M}^{-1}\ve{D})$.  The spectral radius for the Marshak
problem is plotted in Figure~\ref{fig:marshak_spectral}.  As described in
\S~\ref{sec:preconditioning}, the preconditioning effectively reduces the
spectral radius of $\ve{H}$ to less than 1.
\begin{figure}[h!]
  \centerline{ \includegraphics[width=5in,clip]{spectral_radius.pdf}}
  \caption{Spectral radius of the iteration matrix, $\ve{H}$, for the
    Marshak Wave problem as a function of elapsed problem time.}
  \label{fig:marshak_spectral}
\end{figure}

There are two principal degrees of freedom when utilizing the MCSA method: (1)
the number of particles per stage and (2) the weight cutoff.
Figure~\ref{fig:CPU_Np} shows the CPU time as a function of the requested
number of particles per stage for a relative weight cutoff of $\sn{1}{-4}$.
For less than five particles per stage, the method does not converge.
Figure~\ref{fig:CPU_Ns} shows the variation in the maximum number of stages
(iterations) per cycle with number of particles.  Analysis of these two
figures shows that increasing the number of particles per stage reduces the
number of iterations required to converge the solution.  However, the cost of
the Monte Carlo transport is high; thus, better performance is attained by
running more stages with fewer particles per stage \cite{evans_2003}.
\begin{figure}[h!]
  \centerline{ \includegraphics[width=5in,clip]{mrshk_np_CPU.pdf}}
  \caption{CPU time versus number of particles per stage for the
    Marshak problem.  The number of particles is the requested number.
    In each stage the exact number can vary depending upon the number
    of cells and source strength per cell.}
  \label{fig:CPU_Np}
\end{figure}

\begin{figure}[h!]
  \centerline{ \includegraphics[width=5in,clip]{mrshk_np_Ns.pdf}}
  \caption{ Maximum number of stages per cycle versus number of
    particles per stage for the Marshak problem.  The number of
    particles is the requested number.  In each stage the exact number
    can vary depending upon the number of cells and source strength
    per cell.}
  \label{fig:CPU_Ns}
\end{figure}

Figure~\ref{fig:CPU_wc} shows the CPU time as a function of relative weight
cutoff.  These results indicate that the performance of the MCSA method is
relatively insensitive to the weight cutoff.  This is especially true when
running small numbers of particles per stage.  The weight cutoff has a more
pronounced effect on the efficiency of the method when larger numbers of
particles are used.
\begin{figure}[h!]
  \centerline{ \includegraphics[width=5in,clip]{mrshk_wc_CPU.pdf}}
  \caption{ CPU time versus relative weight cutoff for the Marshak
    problem.  Results are shown for 10 and 50 requested particles per
    stage.  The numbers in parentheses show the maximum number of
    particles run in a stage during the course of the problem.}
  \label{fig:CPU_wc}
\end{figure}

Having analyzed the variation of the MCSA algorithm efficiency with number of
particles and weight cutoff, we now compare to traditional deterministic
algorithms.  Table~\ref{tab:marshak_comparison} shows timing results for the
Marshak problem using preconditioned CG (PCG), preconditioned Richardson
iteration (PFIX), and MCSA. The results show that the MCSA method is the most
efficient method, although the PCG results are nearly as fast.  More
importantly, the MCSA method clearly accelerates Richardson iteration, as
indicated by the fact that MCSA is 42\% faster than Richardson iteration
alone.
\begin{table}[h!]
  \caption{Comparison of solution methods for the Marshak problem. All
    methods were run with a stopping criterion of $\sn{1}{-8}$.  The
    MCSA method used $N_p=10$ and $W_c = \sn{1}{-4}$.  The CPU time is
    measured relative to the MCSA method, which required 72.6 s on a
    3.6~GHz Xeon.}
  \label{tab:marshak_comparison}
  \begin{center}
    \begin{tabular}{lrr}\hline\hline
      \multicolumn{1}{c}{Method} & \multicolumn{1}{c}{Max Iterations}
      & \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline  PCG &
      11 & 1.11 \\  PFIX & 11 & 1.42 \\  MCSA & 11 & 1.0
      \\ \hline\hline
    \end{tabular}
  \end{center}
\end{table}

>>>>

A timing and convergence study is used to demonstrate the
effectiveness of the adjoint method as compared to the direct
method. To assess both the CPU time and number of iterations required
to converge to a solution, a problem of constant $\Delta$ was used
with varying values of grid size, fixing the spectral radius of the
system at a constant value for each variation. Both the five-point and
nine-point stencils were used with both the direct and adjoint
solvers. For each case, $N \times N$ total random walk permutations
were computed per MCSA iteration where $N \times N$ is the number of
discrete grid points in the system. Solver parameters were set to a
weight cutoff of \sn{1}{-4} for the stochastic linear solver and a
convergence tolerance of \sn{1}{-8} for the MCSA iterative solver. All
computations presented in this section were completed on a 2.66 GHz
Intel Core i7 machine with 8 GB 1067 MHz DDR3
memory. Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed
for each case to converge in seconds, and
Figure~\ref{fig:poisson_iterations} gives the number of iterations
needed for each case to converge relative to the problem size.
\begin{figure}[ht!]
  \centering
  \includegraphics[width=5in,clip]{dir_adj_cpu.pdf}
  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
    $N \times N$ square mesh). Both the adjoint and direct solvers are
    used with the five point and nine point stencils. A CPU time
    speedup is noted with the adjoint method due to the higher density
    of random walk events in regions with a large residual.}
  \label{fig:poisson_cpu_time}
\end{figure}

\begin{figure}[ht!]
  \centering
  \includegraphics[width=5in,clip]{dir_adj_iterations.pdf}
  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
    \times N$ square mesh). Both the adjoint and direct solvers are
    used with the five-point and nine-point stencils. }
  \label{fig:poisson_iterations}
\end{figure}

@@

As an additional comparison, the convergence behavior of MCSA can be
analyzed using both the adjoint and direct solvers to detect any
performance benefits. To assess the convergence properties of MCSA
using each solver and stencil, the infinity norm of the residual
computed in Eq.~(\ref{eq:MCSA-residual}) was collected at each
iteration for a fixed problem size of
$N=500$. Figure~\ref{fig:poisson_convergence} gives the results of
these computations. First, it is worthy to note on the semilog plot
that we are indeed achieving the expected exponential convergence from
MCSA with both Monte Carlo solvers. Second, we note that using the
adjoint method with the same number of stochastic histores per MCSA
iteration gives a faster rate of converge for the same reasons as
above.
\begin{figure}[ht!]
  \centering
  \includegraphics[width=5in,clip]{dir_adj_conv.pdf}
  \caption{\sl Infinity norm of the solution residual vs. iteration
    number for a problem of size $N=500$. Both the adjoint and direct
    solvers are used with the five point and nine point stencils. A
    higher rate of convergence is observed for MCSA using the adjoint
    Monte Carlo solver as compared to the direct method when both
    solvers compute the same number of random walks per iteration.}
  \label{fig:poisson_convergence}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
  \includegraphics[width=5in,clip]{seq_cpu.pdf}
  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
    $N \times N$ square mesh). Both the Sequential Monte Carlo and
    MCSA solvers are used with the five point stencils and the adjoint
    Monte Carlo solver. The number of random walks was twice the
    number of discrete states in the system in order to ensure
    convergence in the Sequential Monte Carlo method.}
  \label{fig:seq_cpu_time}
\end{figure}

\begin{figure}[ht!]
  \centering
  \includegraphics[width=5in,clip]{seq_iterations.pdf}
  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
    \times N$ square mesh). Both the Sequential Monte Carlo and MCSA
    solvers are used with the five point stencils and the adjoint
    Monte Carlo solver.}
  \label{fig:seq_iterations}
\end{figure}

The benefits of using a synthetic acceleration scheme are also noted
when the infinity norm of the residual computed at each iteration for
both methods was collected at each iteration for a fixed problem sizes
of $N=100$ and $N=500$ as shown in figures Figure~\ref{fig:seq_100}
and \ref{fig:seq_500} respectively. In both cases, the Sequential
method is subject to two regimes of exponential convergence with the
later regime converging the slowest while the MCSA method exhibits a
single rate of exponential convergence, much higher than that computed
by Halton's method. Even with the doubling of the number of stochastic
histories computed per time step in order to ensure convergence for
the Sequential method, we still see robustness issues with a
non-monotonically decreasing residual observed for the $N=100$
case. In both cases the MCSA solver is observed to be robust with a
monotonically decreasing residual.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=5in,clip]{seq_conv_100.pdf}
  \caption{\sl Infinity norm of the solution residual vs. iteration
    number for a problem of size $N=100$. Both the Sequential Monte
    Carlo and MCSA solvers are used with the five point stencils and
    the adjoint Monte Carlo solver.}
  \label{fig:seq_100}
\end{figure}

\begin{figure}[ht!]
  \centering
  \includegraphics[width=5in,clip]{seq_conv_500.pdf}
  \caption{\sl Infinity norm of the solution residual vs. iteration
    number for a problem of size $N=500$. Both the Sequential Monte
    Carlo and MCSA solvers are used with the five point stencils and
    the adjoint Monte Carlo solver.}
  \label{fig:seq_500}
\end{figure}

@@

Applying backward-Euler time differencing to Eq.~(\ref{eq:eq-diff}) and
lagging the temperature-dependent coefficients at $t^n$ gives
\begin{equation}
  -\nabla\cdot \Dn\grad{\phi} + \sign\phi = \qn\:,
\end{equation}
where we have suppressed the $n+1$ indices on $\phi$ and
\begin{align}
  \sign &= \frac{\rho \Cv^n}{4ac(\Tn)^3\dt} + \frac{1}{c\dt}\:,\\ \qn
  &= \sign\phin + Q^n\:,\\ \phin &= ac(\Tn)^4\:.
\end{align}
Applying Fick's Law,
\begin{equation}
  \ve{F} = -D\grad{\phi}\:,
  \label{eq:fick}
\end{equation}
where $\ve{F}$ is the radiation flux, we define a flux-balance equation,
\begin{equation}
  \dive{F} + \sign\phi = \qn\:.
  \label{eq:flux-balance}
\end{equation}
Integrating Eq.~(\ref{eq:flux-balance}) over the cell illustrated in
Figure~\ref{fig:cell} gives
\begin{equation}
  \iiint\dive{F}\,d\ve{r} +
  \sign_{ijk}\phi_{ijk}V_{ijk}=\qn_{ijk}V_{ijk}\:,
  \label{eq:flux-balance-vol-integral}
\end{equation}
and $V_{ijk} = \Delta_i\Delta_j\Delta_k$.
\begin{figure}[ht!]
  \centerline{ \includegraphics[clip]{mesh_cell.pdf} }
  \caption{Orthogonal 3D mesh cell.}
  \label{fig:cell}
\end{figure}
We now apply the divergence theorem,
\begin{equation}
  \begin{split}
    \iiint\dive{F}\,d\ve{r} &= \oint\hat{\ve{n}}\cdot\ve{F}\,dA\\ &=
    \sum_{f=1}^{6}\hat{\ve{n}}_f\cdot\ve{F}_f A_f\:,
  \end{split}
\end{equation}
to Eq.(\ref{eq:flux-balance-vol-integral}) yielding
\begin{equation}
  \begin{split}
    (F_{i+1/2} - F_{i-1/2})\Dj\Dk &+ (F_{j+1/2} - F_{j-1/2})\Di\Dk +
    (F_{k+1/2}\\ &- F_{k-1/2})\Di\Dj + \sign_{ijk}\phi_{ijk}V_{ijk} =
    \qn_{ijk}V_{ijk}\:.
  \end{split}
  \label{eq:flux-balance-difference}
\end{equation}

@@

The connectivity of an arbitrary cell is illustrated in the 3D mesh segment in
Figure~\ref{fig:mesh_segment}.  For a particle residing in the blue cell, any
collision results in a scatter to one of the six adjacent cells since
$p_{ii}=0$ in the preconditioned system.  We can easily calculate the
probabilities for each scattering event from
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) without explicitly forming
the iteration matrix.  In this mesh cell, the probabilities for leakage out of
each face are derived from the equations from each adjacent cell.  Using
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) we can explicitly write the
equations from each adjacent cell, here only noting the contribution from cell
7:
\begin{align*}
  \phi_1 &= \frac{\sigm_{7\rightarrow 1}}{\sigma_{T\,1}}
  \frac{\Delta_7}{\Delta_1}\phi_7 + \ldots\:,\\ 
  %%
  \phi_2 &= \frac{\sigp_{7\rightarrow 2}}
  {\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2}\phi_7 + \ldots\:,\\
  %%
  \phi_3 &= \frac{\sigm_{7\rightarrow 3}}{\sigma_{T\,3}}
  \frac{\Delta_7}{\Delta_3}\phi_7 + \ldots\:,\\ 
  %%
  \phi_4 &= \frac{\sigp_{7\rightarrow 4}}{\sigma_{T\,4}}
  \frac{\Delta_7}{\Delta_4}\phi_7 + \ldots\:,\\
  %%
  \phi_5 &= \frac{\sigm_{7\rightarrow 5}}{\sigma_{T\,5}}
  \frac{\Delta_7}{\Delta_5}\phi_7 + \ldots\:,\\
  %%
  \phi_6 &= \frac{\sigp_{7\rightarrow 6}}{\sigma_{T\,6}}
  \frac{\Delta_7}{\Delta_6}\phi_7 + \ldots\:.\\ 
\end{align*} 
A quick view of these equations reveals that the only non-zero entries in
column 7 of the iteration matrix are
\begin{equation*}
  \begin{pmatrix}
    h_{1,7} & h_{2,7} & h_{3,7} & h_{4,7} & h_{5,7} & h_{6,7}
  \end{pmatrix}^{T}
  =
  \begin{pmatrix}
    \frac{\sigm_{7\rightarrow
        1}}{\sigma_{T\,1}}\frac{\Delta_7}{\Delta_1}
    \\ \frac{\sigp_{7\rightarrow
        2}}{\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2}
    \\ \frac{\sigm_{7\rightarrow
        3}}{\sigma_{T\,3}}\frac{\Delta_7}{\Delta_3}
    \\ \frac{\sigp_{7\rightarrow
        4}}{\sigma_{T\,4}}\frac{\Delta_7}{\Delta_4}
    \\ \frac{\sigm_{7\rightarrow
        5}}{\sigma_{T\,5}}\frac{\Delta_7}{\Delta_5}
    \\ \frac{\sigp_{7\rightarrow
        6}}{\sigma_{T\,6}}\frac{\Delta_7}{\Delta_6}
  \end{pmatrix}\:,
\end{equation*}
where, following Eq.~(\ref{eq:point_iteration}), $\ve{H} = (\vI -
\ve{M}^{-1}\ve{D})$ is the iteration matrix.  Now, using
Eq.~(\ref{eq:adjoint-probability}), we define the scattering probabilities for
cell 7 as
\begin{equation*}
  P_{7\rightarrow n} = p_{7,n} = \frac{|h_{n,7}|}{|h_{1,7} + h_{2,7} +
    h_{3,7} + h_{4,7} + h_{5,7} + h_{6,7}|}\:,\quad n\in
  \{1,\ldots,6\}\:.
\end{equation*}
The weight change for each scattering event is calculated using
Eq.~(\ref{eq:adjoint-weight}):
\begin{equation*}
  W_{m} = \frac{h_{n,7}}{p_{7,n}}W_{m-1}\:,\quad n\in
  \{1,\ldots,6\}\:,
\end{equation*}
where $m$ is the step-index in the particle history.  We also note that, for
the 3D Cartesian mesh considered in this work, the maximum number of non-zero
entries in columns of the iteration matrix is 6, generating a sparse system.
Cells along the problem boundary can have fewer than six entries.
\begin{figure}[ht!]
  \centerline{ \includegraphics[clip]{3D_mesh_segment.pdf}}
  \caption{3D mesh-segment.  The particle's current cell (in blue) is
    connected to 6 adjacent cells.  The cell numbers are plotted
    alongside the mesh such that the blue cell has index 7, the high
    $z$ cell has index 2, etc.}
  \label{fig:mesh_segment}
\end{figure}

@@

Figure~\ref{fig:marshak_10} shows the analytic Marshak Wave solution compared
to the MCSA solution.  As opposed to most Monte Carlo techniques, the MCSA
method generates solutions that are numerically equivalent to the PCG and PFIX
results.  Because the MCSA method converges the residual of the system, the
numerical accuracy is determined by the stopping criterion shown in
Eq.~(\ref{eq:stopping-criteria}).  Also, the MCSA method is not bound by the
Central Limit Theorem so convergence is not restricted to
$1/\sqrt{N}$~\cite{halton_1994,evans_2003}.
\begin{figure}[ht!]
  \centerline{ \includegraphics[width=5in,clip]{marshak_10.pdf}}
  \caption{ Solution of the Marshak wave problem at $t=100$~ns.  The analytic
    solution is compared to the MCSA solution.  The MCSA, PCG, and PFIXED
    methods are all converged to the same tolerance and therefore give
    numerically equivalent answers.}
  \label{fig:marshak_10}
\end{figure}

As stated in \S~\ref{sec:monte-carlo-matrix}, the MCSA method will converge
when $\rho(\ve{H})< 1$, where $\ve{H}$ is the iteration matrix and is defined
$\ve{H} = (\vI - \ve{M}^{-1}\ve{D})$.  The spectral radius for the Marshak
problem is plotted in Figure~\ref{fig:marshak_spectral}.  As described in
\S~\ref{sec:preconditioning}, the preconditioning effectively reduces the
spectral radius of $\ve{H}$ to less than 1.
\begin{figure}[ht!]
  \centerline{ \includegraphics[width=5in,clip]{spectral_radius.pdf}}
  \caption{Spectral radius of the iteration matrix, $\ve{H}$, for the
    Marshak Wave problem as a function of elapsed problem time.}
  \label{fig:marshak_spectral}
\end{figure}

There are two principal degrees of freedom when utilizing the MCSA method: (1)
the number of particles per stage and (2) the weight cutoff.
Figure~\ref{fig:CPU_Np} shows the CPU time as a function of the requested
number of particles per stage for a relative weight cutoff of $\sn{1}{-4}$.
For less than five particles per stage, the method does not converge.
Figure~\ref{fig:CPU_Ns} shows the variation in the maximum number of stages
(iterations) per cycle with number of particles.  Analysis of these two
figures shows that increasing the number of particles per stage reduces the
number of iterations required to converge the solution.  However, the cost of
the Monte Carlo transport is high; thus, better performance is attained by
running more stages with fewer particles per stage \cite{evans_2003}.
\begin{figure}[ht!]
  \centerline{ \includegraphics[width=5in,clip]{mrshk_np_CPU.pdf}}
  \caption{CPU time versus number of particles per stage for the
    Marshak problem.  The number of particles is the requested number.
    In each stage the exact number can vary depending upon the number
    of cells and source strength per cell.}
  \label{fig:CPU_Np}
\end{figure}

\begin{figure}[ht!]
  \centerline{ \includegraphics[width=5in,clip]{mrshk_np_Ns.pdf}}
  \caption{ Maximum number of stages per cycle versus number of
    particles per stage for the Marshak problem.  The number of
    particles is the requested number.  In each stage the exact number
    can vary depending upon the number of cells and source strength
    per cell.}
  \label{fig:CPU_Ns}
\end{figure}

Figure~\ref{fig:CPU_wc} shows the CPU time as a function of relative weight
cutoff.  These results indicate that the performance of the MCSA method is
relatively insensitive to the weight cutoff.  This is especially true when
running small numbers of particles per stage.  The weight cutoff has a more
pronounced effect on the efficiency of the method when larger numbers of
particles are used.
\begin{figure}[ht!]
  \centerline{ \includegraphics[width=5in,clip]{mrshk_wc_CPU.pdf}}
  \caption{ CPU time versus relative weight cutoff for the Marshak
    problem.  Results are shown for 10 and 50 requested particles per
    stage.  The numbers in parentheses show the maximum number of
    particles run in a stage during the course of the problem.}
  \label{fig:CPU_wc}
\end{figure}

Having analyzed the variation of the MCSA algorithm efficiency with number of
particles and weight cutoff, we now compare to traditional deterministic
algorithms.  Table~\ref{tab:marshak_comparison} shows timing results for the
Marshak problem using preconditioned CG (PCG), preconditioned Richardson
iteration (PFIX), and MCSA. The results show that the MCSA method is the most
efficient method, although the PCG results are nearly as fast.  More
importantly, the MCSA method clearly accelerates Richardson iteration, as
indicated by the fact that MCSA is 42\% faster than Richardson iteration
alone.
\begin{table}[ht!]
  \caption{Comparison of solution methods for the Marshak problem. All
    methods were run with a stopping criterion of $\sn{1}{-8}$.  The
    MCSA method used $N_p=10$ and $W_c = \sn{1}{-4}$.  The CPU time is
    measured relative to the MCSA method, which required 72.6 s on a
    3.6~GHz Xeon.}
  \label{tab:marshak_comparison}
  \begin{center}
    \begin{tabular}{lrr}\hline\hline
      \multicolumn{1}{c}{Method} & \multicolumn{1}{c}{Max Iterations}
      & \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline  PCG &
      11 & 1.11 \\  PFIX & 11 & 1.42 \\  MCSA & 11 & 1.0
      \\ \hline\hline
    \end{tabular}
  \end{center}
\end{table}
